---
title: "Study title"
author: "Your name, supervisor 1, supervisor 2, etc"
date: "Today's date"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: true
    theme: cerulean
    highlight: pygments
editor_options:
  chunk_output_type: console
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding=encoding, output_file=file.path(dirname(inputFile), 'tutorial.html')) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=TRUE)
```

**Note:** This template includes some basic structure to help you build your analyses, but won't run since it doesn't know what your data looks like yet!

# Background

*Write something here to introduce your project, your dataset, and how you plan to analyse it.*

# Setup

We used the [pacman Package Management Tool](https://cran.r-project.org/web/packages/pacman/index.html) to install and load subsequent packages in a condensed and efficient way.

```{r, eval=FALSE}
# Install pacman (do once per computer)
install.packages("pacman")
```

```{r, results='hide', warning=FALSE, message=FALSE}
# Install and load required packages
pacman::p_load(ggplot2, ggpubr, janitor, lme4, lubridate, readr, readxl, sf, sp, tidyverse, viridis)
```

# Data preparation

First, we read in our data.

```{r,  warning=FALSE, message=FALSE}
# Read in data
data <- read_excel("input/raw data.xlsx", sheet="Sheet1") %>%
  # Change column names so they are in lower case and multiple words are
  # separated with underscores
  clean_names() %>%
  # Create a date column and tell R to recognise it as a date
  mutate(date=as.Date(start_time))

# Save the processed data as a new file
write.csv(data, "output/processed data.csv")
```

Next, we explore the data.

```{r}
# Look at first 6 rows of the data
head(data)

# Look at last 6 rows of the data
tail(data)

# Look at the structure of the data
str(data)

# Look at the column names
names(data)

# Look at unique values for a single column
unique(data$site)

# Find out how many rows there are in the data
nrow(data)
```

We plotted our continuous data.

```{r}
# Map points
plot <- ggplot(data=data, aes(x=column_1, y=column_2)) + 
  # Add the spatial features (points) to the plot
  geom_smooth() +
  # Remove grey background
  theme_minimal() + 
  # Use red-green colourblind friendly colours
  scale_colour_manual(values=viridis()) + 
  # Add x and y axis labels
  xlab("Column 1") + ylab(" Column 2")

# Display the plot
print(plot)
```

```{r, eval=FALSE, include=FALSE}
# Save the plot as a jpeg
ggsave(plot=plot, "output/plot.jpeg", 
       height=6, width=12, dpi=300)
```

# Models

Here we test our hypotheses by fitting generalised linear models (GLMs). The benefit of GLMs is they can deal with such categorical predictors, are relatively easy to interpret, and are robust against overfitting. However, they are sensitive to outliers, and can’t deal with presence-only data.

```{r}
# Fit a GLM
mod <- glm(response_1 ~ predictor_2 + predictor_3, data=data)

# Display summary statistics
summary(mod)
```

If the `Pr(>|t|)` (or *p*-value) for a predictor is <0.05, it *significantly* affected the response. 

# Plots

Here we plot our significant results.

```{r}
# Plot 
```

# Session information

```{r}
# Display version information for R, OS, and packages
sessionInfo()
```